Autopolyploid establishment through polygenic adaptation

### 2024-12-17 doc/paper/paper.tex
update

\documentclass[11pt,a4paper]{article}
\input{preamble.tex}
\usepackage{geometry}
\usepackage{authblk}
\externaldocument[s-]{supp}

\begin{document}
\title{Autopolyploid establishment through polygenic adaptation}
\author[1]{Arthur Zwaenepoel\thanks{arthur.zwaenepoel@univ-lille.fr}}
\affil[1]{\footnotesize University of Lille, CNRS, UMR 8198 -- Evo-Eco-Paleo,
F-59000 Lille, France}
\date{\vspace{-5ex}}
\maketitle
\begin{abstract}
We define the infinitesimal model of quantitative genetics (\textit{sensu}
\cite{barton2017}) for the inheritance of an additive quantitative trait in a
mixed-ploidy population consisting of diploid, triploid and autotetraploid
individuals producing haploid and diploid gametes.
We implement efficient simulation methods and use these to study the
quantitative genetics of mixed-ploidy populations and the establishment of
autotetraploids in a new habitat.
We show that in the absence of migration from the source population,
autotetraploids are more likely to found a successful population than diploids
under a very broad range of conditions, but that this is unlikely to
sufficiently counter the scarcity of tetraploid founders when the source is
predominantly diploid.
[...]
\end{abstract}

@@

@@

Consider for instance a randomly mating diploid population.
Under the commonly accepted view that polyploids mostly emerge through the
union of unreduced gametes \citep{herben2016,kreiner2017b}, a new tetraploid
individual originating by a chance encounter of two unreduced diploid gametes
(an event occurring at an appreciable rate, \citep{kreiner2017}) is highly
unlikely to establish a stable tetraploid subpopulation, as most of its gametes
will end up in unfit hybrids of odd ploidy level (triploids).
This negative frequency dependence effect is commonly referred to as
\textit{minority cytotype exclusion} (MCE), after \cite{levin1975}.
It is well-appreciated that, due to MCE, in a large random mating mixed-ploidy
population dominated by diploids, the rate of unreduced gamete formation needs
to be extraordinarily high for tetraploids to establish (\cite{felber1997}, see
also \cref{s-sec:det}).

Hence, to explain the widespread occurrence of mixed-ploidy populations,
additional factors besides the continuous formation of polyploids through the
union of unreduced gametes need to be considered.
On the one hand, chance establishment of tetraploids through drift could occur.
However, MCE is quite strong in randomly mating mixed-ploidy populations, and
the population size has to be very small for tetraploid establishment to occur
at an appreciable rate (\cite{rausch2005}, see also \cref{s-sec:mchain}).
On the other hand, any form of prezygotic isolation between cytotypes could
promote establishment of polyploid cytotypes by alleviating MCE.
Particularly relevant are assortative mating by cytotype  (for instance through
phenological differences across cytotypes, or differences in pollinators,
\citep{kolar2017}), self-fertilization \citep{rausch2005}, and localized
dispersal \citep{baack2005,kolar2017}.
Finally, selection may be invoked to explain the establishment of polyploids.
Tetraploids may have higher relative fitness than their diploid counterparts
due to reduced inbreeding depression \citep{ronfort1999}, or due to being
better adapted to (changing) environmental conditions \citep{vandepeer2021}.

None of these factors is likely to explain by itself the establishment of
polyploids, and the consensus in the field appears to be that some mix of the
above is required to explain the occurrence of mixed-ploidy populations in
nature \citep{kolar2017,mortier2024}.
In particular, polyploids are thought to establish mainly in novel, unoccupied
habitats where they evade MCE (for instance at range edges, or after local
extinction due to environmental change).
If they are able to colonize such habitats at an appreciable rate, they must
somehow be better adapted to local conditions, or more able to adapt to those
conditions despite inbreeding and migration from the source population while
the population is still small, compared to diploids.
Indeed, when the source population is dominated by diploids, the number of
tetraploid migrants arriving in the novel habitat will be almost negligible
($O(u^2)$ if $u$ is the probability that a diploid produces an unreduced
gamete in meiosis) compared to the number of diploid migrant individuals, so
that tetraploids need a substantial advantage during colonization if they are
to establish with reasonable probability.

While there have been substantial modeling efforts aimed at understanding
autotetraploid establishment within diploid populations (e.g. \cite{levin1975,
felber1991, felber1997, rausch2005, oswald2011, clo2022c, griswold2021}), the
problem of polyploid establishment in a novel habitat presenting some adaptive
challenge remains largely unaddressed, despite its centrality to verbal
arguments about the establishment of polyploids in natural populations
\citep{kolar2017, vandepeer2021, clo2022d}.

Here we develop a model for the establishment of a mixed-ploidy population in a
novel, unoccupied habitat based on \cite{barton2018}.
In order to establish in the novel habitat, the population has to adapt to
local environmental conditions.
We assume fitness is determined by directional selection on a single polygenic
trait which can be interpreted as log fitness at low density in the new
habitat.
As in \cite{barton2018}, we assume the trait follows the infinitesimal
model (\textit{sensu} \cite{barton2017}, i.e. the `Gaussian descendants'
infinitesimal model).
We extend the infinitesimal model, and the approach for exact simulation of
trait evolution under the infinitesimal model, to mixed-ploidy populations.
We then use simulations to study tetraploid establishment, both from single
migrants and under continuous migration from a predominantly diploid source
population, examining the effects of polyploid (quantitative) genetics,
maladaptive migration, selfing and assortative mating on the probability that
autotetraploids establish in the novel habitat before diploids do.

@@

\begin{table}[t]
\caption{Glossary of the main notation used in the main text.
} \label{tbl:glossary}
\centering
\small
\begin{tabularx}{\linewidth}{lX}
\cline{1-2}
\textbf{notation}   & \textbf{description}   \\ \cline{1-2}
$N, N_k$ & total population size, population size of the $k$-ploid cytotype \\
$\pi_k$ & equilibrium frequency of the $k$-ploid cytotype \\
$u$ & probability of unreduced gamete formation ($u=u_{22}=u_{32}=1-u_{42}$)\\
$v$ & probability that a triploid produces a haploid/diploid gamete
  ($v=u_{31}=u_{32}$)\\
$m$ & expected number of migrants per generation arriving in the new habitat \\
$z_i$ & trait value of individual $i$ \\
$c_i$ & ploidy level of individual $i$ \\
$g_i$ & ploidy level of gamete produced by individual $i$ in a particular cross\\
$V$ & segregation variance in the reference diploid population \\
$V_{i,k}$ & gametic segregation variance associated with the production of a
  $k$-ploid gamete by individual $i$ \\
$\Vx_k$ & genetic variance associated with a haploid genome in the $k$-ploid
  reference population (i.e. a $k$-ploid non-inbred population at HWLE) \\
$\beta_{k}$ & scaling factor for allelic effects in $k$-ploids \\
$F_i$ & inbreeding coefficient in individual $i$ \\
$\Phi_{ij}$ & coancestry coefficient for individuals $i$ and $j$ \\
$\alpha_k$ & probability that the two genes at a locus in a diploid gamete
  formed by a $k$-ploid individual descend from the same parental gene copy\\
$\gamma$ & strength of directional selection in the new habitat\\
$\theta$ & trait value beyond which the growth rate becomes positive in the new
habitat \\ $w_{ij}$ & fitness of parental pair $(i,j)$ \\
$w_{ij}^{kl}$ & expected fitness of offspring from parental pair $(i,j)$ when
$i$ contributes a $k$-ploid gamete and $j$ contributes a $l$-ploid gamete \\
$\sigma_k$ & selfing rate in $k$-ploids \\
$\rho_k$ & probability of assortative mating in $k$-ploids \\
\cline{1-2}
\end{tabularx}%
\end{table}

@@

When two individuals mate, we assume they produce gametes according to their
ploidy level (\cref{eq:U}), which randomly combine to produce offspring (which
may be inviable if one of the contributing gametes is inviable).
Intrinsic fitness disadvantages associated with particular zygotic ploidy
levels or cross types (e.g. modeling phenomena such as `triploid block',
\citep{ramsey1998,brown2024}) can be straighforwardly included at this level.
An analysis of a deterministic model (i.e. where $N \rightarrow \infty$) for
the cytotype dynamics and equilibrium cytotype composition under random mating
is included in \cref{s-sec:det} (see also \cite{felber1997,kauai2024}).
The stochastic version for finite and constant $N$ is analyzed briefly in
\cref{s-sec:mchain}.

@@

We model partial self-fertilization by assuming that a proportion
$\sigma_{c_i}$ of the ovules of individual $i$ with ploidy level $c_i$ are
fertilized by self-pollen, while the remaining proportion $1-\sigma_{c_i}$ are
fertilized by randomly sampled pollen (which may be self-pollen with
probability $1/N$). 
That is, the expected number of offspring from individual $i$ as mother
surviving after selection is
\begin{equation}
\Ex[w_i] = \sigma_{c_i} \Ex[w_{ii}] +
  (1-\sigma_{c_i})\left[\frac{1}{N}\sum_{j=1}^N \Ex[w_{ij}]\right]
\end{equation}
Where $\Ex[w_{ij}] = e^{\gamma \bar{z}_{ij} + \gamma^2 V_{ij}/2}$ is the
expected fitness of offspring from the parental pair with $i$ as mother and $j$
as father.
We hence assume no pollen limitation (all outcrossing ovules are fertilized),
and no pollen discounting (the probability of being a father is unaffected by
an individual's selfing rate).
When modeling self-incompatibility, we assume there is no intrinsic
disadvantage to self-incompatibility, except when there is only a single
individual in the population i.e.
\begin{equation}
\Ex[w_i] = \begin{cases}
    \frac{1}{N-1}\sum_{j \ne i }\Ex[w_{ij}] & \text{if } N > 1\\ 
    0 & \text{if } N=1 \end{cases}
\end{equation}

We model assortative mating by ploidy level in a similar way, assuming that a
fraction $\rho_{c_i}$ of the ovules of individual $i$ are fertilized by pollen
sampled from the $c_i$-ploid portion of the population, while a fraction
$1-\rho_{c_i}$ is fertilized by pollen randomly sampled from the entire
population.
\begin{equation}
\Ex[w_i] = \sum_{j=1}^N \left(\frac{\delta_{c_i,c_j}}{N_{c_i}}\rho_{c_i} +
\frac{1}{N}(1-\rho_{c_i})\right)\Ex[w_{ij}]
\end{equation}

@@

Simulations for the mixed-ploidy model further confirm the correctness of our
infinitesimal approximation and highlight the importance of assumptions
regarding the scaling of allelic effects across ploidy levels
($\beta$ parameters, \cref{s-fig:vz}).
Although inbreeding is slower in autotetraploids than in diploids for the same
population size, the tetraploid fraction of a diploid-dominated mixed-ploidy 
population will have an equal or higher average inbreeding coefficient
(\cref{s-fig:f1}).
This is because in such a population, triploid and tetraploid individuals
mostly arise from gametes formed by diploid individuals or polyploid
individuals with very recent diploid ancestry (about $1+u+2v$ generations ago
for tetraploids, and $1+\frac{2}{3}(u+2v)$ generations ago for triploids, see
\cref{s-sec:ttdip}), so that the polyploid subpopulations will show an average
relatedness similar to the diploid population and not evolve as an isolated
higher-ploidy population would.
A nonzero probability of producing IBD diploid gametes ($\alpha_k > 0$) will
then further increase the inbreeding coefficient in the tetraploid and triploid
fraction of the population relative to their diploid progenitors
(\cref{s-fig:f1}).
Therefore, as long as diploids dominate, the effect of harboring some of the
gene pool in triploid and tetraploid individuals on the rate of inbreeding is
negligible, and we find that the evolution of the inbreeding coefficient over
time is well predicted by $1-e^{-t/2N_e}$, where the inbreeding-effective
population size is, to first order in $u$, given by $(1-2u)N$
(\cref{s-sec:effsize}).

@@

@@

We first consider the establishment of a population from a single migrant
individual with trait value $z_0 = 0$.
We assume $u=0$ and compare the probability of establishment when the migrant
is diploid vs. tetraploid (\cref{fig:est1}).
As noted by \cite{barton2018}, the establishment probability depends
essentially on two dimensionless parameters, $\gamma\sqrt{2V}$ and
$\theta/\sqrt{2V}$, corresponding to the intensity of selection and the degree of
maladaptation, respectively.
We shall scale our results accordingly, assuming $2V=1$ throughout.

We find that reduced inbreeding in tetraploids substantially increases the
establishment probability.
Indeed, in the case where allelic effects are scaled so as to yield the same
equilibrium genetic variance ($\beta = \sqrt{1/2}$), the establishment probability 
for tetraploids can be almost five times as high as in diploids depending on
the selection gradient (\cref{fig:est1}A).
While the probability of establishment becomes lower as the strength of
selection becomes large, the establishment probability does not decrease
monotonically with $\gamma$, i.e. as selection becomes very weak, the
establishment probability also lowers.
This happens because, although the probability of surviving the first couple of
generations becomes higher for weaker selection, adaptation will be slower, and
the risk that genetic variation is exhausted due to inbreeding before the
population is able to reach a consistently positive growth rate is increased.

As expected, the scaling of the equilibrium genetic variance has a profound
effect on the establishment probability, but only when $\beta$ is close to 0.5
(i.e. individual alleles have almost half the effect size in tetraploids
compared to diploids) is the benefit of the slower rate of inbreeding in
tetraploids canceled (\cref{fig:est1}B).

\begin{figure}[t]
\centering
\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/est1b.pdf}
\caption{
(A) Probability of establishment from a single diploid or tetraploid individual
with trait value $z=0$ for increasing selection intensity $\gamma$.
We assume $m=0$ and $u=0$, i.e. there is no migration, and no unreduced
gametes are produced. The trait is scaled in tetraploids so as to yield the
same genetic variance at HWLE ($\beta_4 = \sqrt{1/2}$)
(B) Probability of a tetraploid individual with trait value $z = 0$
successfully founding a population ($P_4$), relative to the probability for a
diploid individual with the same trait value ($P_2$).
The vertical dashed line marks $\beta_4=\sqrt{1/2}$, for which the variance at
HWLE is identical between diploids and autotetraploids.
All results are estimated from 500.000 replicate simulations.
\label{fig:est1}}
\end{figure}

@@

We next consider establishment in the new habitat when there is a continuous
influx of migrants coming from a large diploid-dominated mixed-ploidy source
population at cytotype equilibrium.
In this case, establishment is certain to happen eventually, and we are
interested in the probability that a tetraploid population established before a
diploid one does.

We hypothesized that two counteracting processes affect the probability of
autotetraploid establishment in this case.
On the one hand, increased migration will increase the probability that an
otherwise likely successful tetraploid migrant suffers from MCE in the early
generations while the population size is low, because migrants are likely to be
diploid.
On the other hand, tetraploids are strongly reproductively isolated from
the typical migrant, so that they are less prone to maladaptive swamping.
Hence, conditional on evading MCE, they should be able to adapt to the new
habitat at a rate which is not strongly affected by the migration rate.

\begin{figure}[t]
\centering
\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration1.pdf}
\caption{
Establishment with recurrent migration.
(A) Expected time until a population is established in the novel habitat for
increasing rates of migration and different degrees of maladaptation
($\bar{z}_s$, the mean trait value in the source population). 
(B) Proportion of simulation replicates in which tetraploids established. 
The shaded area corresponds to Jeffreys' 95\% interval.
The horizontal dashed line marks the proportion of tetraploid migrants (i.e.
the proportion of tetraploids at equilibrium in the source population).
All results are based on 10.000 replicate simulations.
We assume $\gamma=0.25$ and $u=v=0.05$.
\label{fig:estwmig}}
\end{figure}

\subsection*{Loss of self-incompatibility}

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration-si.pdf}
\caption{
Establishment with recurrent migration and loss of self-incompatibility in
polyploids.
(A) Expected time until a population is established in the novel habitat for
increasing rates of migration and different self-fertilization rates in
polyploids. Diploids are assumed to be self-incompatible.
Note that $\sigma=0.0$ refers to random self-fertilization (i.e.
self-fertilization with probability $1/N$). 
(B) Proportion of simulation replicates in which tetraploids established. 
The shaded area corresponds to Jeffreys' 95\% interval.
All results are based on 10.000 replicate simulations.
We assume $\gamma=0.25$ and $u=v=0.05$.
\label{fig:}}
\end{figure}

\subsection*{Assortative mating}

\begin{figure}
\includegraphics[width=\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration-am.pdf}
\caption{
\label{fig:}}
\end{figure}

>>>>

\documentclass[12pt,a4paper]{article}
\input{preamble.tex}
\usepackage[margin=1in]{geometry}
\usepackage{authblk}
\externaldocument[s-]{supp}

\begin{document}
\title{Autopolyploid establishment through polygenic adaptation}
\author[1]{Arthur Zwaenepoel\thanks{arthur.zwaenepoel@uantwerpen.be}}
\affil[1]{\footnotesize Department of Biology, University of Antwerp, 2020 Antwerp, Belgium}
\date{\vspace{-5ex}}
\maketitle
\begin{abstract}
We define the infinitesimal model of quantitative genetics (\textit{sensu}
\cite{barton2017}) for the inheritance of an additive quantitative trait in a
mixed-ploidy population consisting of diploid, triploid and autotetraploid
individuals producing haploid and diploid gametes.
We implement efficient simulation methods and use these to study the
quantitative genetics of mixed-ploidy populations and the establishment of
autotetraploids in a new habitat.
We show that in the absence of migration from the source population,
autotetraploids are more likely to found a successful population than diploids
under a very broad range of conditions, but that this is unlikely to
sufficiently counter the scarcity of tetraploid founders when the source is
predominantly diploid.
[...]
\end{abstract}

@@

% Should connect more with literature and concepts of isolating mechanisms pre
% vs. postzygotic isolation, and reinforcement etc. (see e.g. Parvathy &
% Himani's paper introduction)

@@

Consider for instance a randomly mating diploid population.
Under the commonly accepted view that polyploids mostly emerge through the
union of unreduced gametes \citep{bretagnolle1995,herben2016,kreiner2017b}, a
new tetraploid individual originating by a chance encounter of two unreduced
diploid gametes (an event occurring at an appreciable rate; \cite{kreiner2017})
is highly unlikely to establish a stable tetraploid subpopulation, as most of
its gametes will end up in unfit hybrids of odd ploidy level (triploid block,
\cite{ramsey1998,kohler2010,brown2024}).
This negative frequency dependence effect is commonly referred to as
\textit{minority cytotype exclusion} (MCE), after \cite{levin1975}.
It is well-appreciated that, due to MCE, in a large random mating mixed-ploidy
population dominated by diploids, the rate of unreduced gamete formation needs
to be extraordinarily high for tetraploids to establish (\cite{felber1997}, see
also \cref{s-sec:det}).

Hence, to explain the widespread occurrence of mixed-ploidy populations,
additional factors besides the continuous formation of polyploids through the
union of unreduced gametes need to be considered.
Firstly, chance establishment of tetraploids through drift could occur.
Indeed, the problem is somewhat analogous to the spread of underdominant
chromosomal rearrangements, where local establishment through genetic drift and
subsequent spreading in a subdivided population through local extinction and 
recolonization dynamics has been suggested as a plausible model
\citep{lande1985}.
However, MCE is quite strong in randomly mating mixed-ploidy populations, and
the population size has to be very small for local tetraploid establishment to
occur at an appreciable rate (\cite{rausch2005}, see also \cref{s-sec:mchain}).
Secondly, any form of prezygotic isolation between cytotypes could
promote establishment of polyploid cytotypes by alleviating MCE.
Particularly relevant are assortative mating by cytotype  (for instance through
phenological differences across cytotypes, or differences in pollinators;
\cite{kolar2017}), self-fertilization \citep{rausch2005,novikova2023}, and
localized dispersal \citep{baack2005,kolar2017}.
Finally, selection may be invoked to explain the establishment of polyploids.
Tetraploids may have higher relative fitness than their diploid counterparts
due to reduced inbreeding depression \citep{ronfort1999}, or due to being
better adapted to (changing) environmental conditions \citep{vandepeer2021}.

None of these factors is likely to explain by itself the establishment of
polyploids, and the consensus in the field appears to be that some mix of the
above is required to explain the occurrence of mixed-ploidy populations in
nature \citep{kolar2017,mortier2024}.
In particular, polyploids are thought to establish mainly in novel, unoccupied
habitats where they evade MCE (for instance at range edges, or after local
extinction due to environmental change).
However, such peripheral habitats are likely to present adaptive challenges to
establishment \citep{kawecki2008}, and if polyploids are able to colonize such
habitats at an appreciable rate, they must somehow be better adapted to local
conditions, or more able to adapt to those conditions despite inbreeding and
migration from the source population while the population is still small,
compared to diploids.
Indeed, when the source population is dominated by diploids, the probability
that a migrant individual is tetraploid will be very small ($O(u^2)$ if $u$ is
the probability that a diploid in the source population produces an unreduced
gamete in meiosis), so that tetraploids need a substantial advantage during
colonization if they are to establish before diploids do.

While there have been substantial modeling efforts aimed at understanding
autotetraploid establishment within diploid populations (e.g. \cite{levin1975,
felber1991, felber1997, rausch2005, oswald2011, clo2022c, griswold2021}), the
problem of polyploid establishment in a novel habitat which presents some adaptive
challenge remains largely unaddressed, despite its centrality to verbal
arguments about the establishment of polyploids in natural populations
\citep{kolar2017, vandepeer2021, clo2022d}.

Here we develop a model for the establishment of a mixed-ploidy population in a
novel, unoccupied habitat based on \cite{barton2018}.
In order to establish in the novel habitat, the population has to adapt to
local environmental conditions.
We assume fitness is determined by directional selection on a single polygenic
trait which can be interpreted as log fitness at low density in the new
habitat.
As in \cite{barton2018}, we assume the trait follows the infinitesimal
model (\textit{sensu} \cite{barton2017}, i.e. the `Gaussian descendants'
infinitesimal model \citep{turelli2017}).
We extend the infinitesimal model, and the approach for exact simulation of
trait evolution under the infinitesimal model, to mixed-ploidy populations.
We then use simulations to study tetraploid establishment, both from single
migrants and under continuous migration from a predominantly diploid source
population, examining the effects of autopolyploid genetics,
maladaptive migration, selfing and assortative mating on the probability that
autotetraploids establish in the novel habitat.

@@

\begin{table}[t]
\caption{Glossary of the main notation used in the main text.
} \label{tbl:glossary}
\centering
\small
\begin{tabularx}{\linewidth}{lX}
\cline{1-2}
\textbf{notation}   & \textbf{description}   \\ \cline{1-2}
$N$ & total population size\\
$N_k$ & population size of the $k$-ploid cytotype \\
$\pi_k$ & equilibrium frequency of the $k$-ploid cytotype \\
$u$ & probability of unreduced gamete formation ($u=u_{22}=u_{32}=1-u_{42}$)\\
$v$ & probability that a triploid produces a haploid/diploid gamete
  ($v=u_{31}=u_{32}$)\\
$m$ & expected number of migrants per generation arriving in the new habitat \\
$z_i$ & trait value of individual $i$ \\
$c_i$ & ploidy level of individual $i$ \\
$g_i$ & ploidy level of gamete produced by individual $i$ in a particular cross\\
$V$ & segregation variance in the reference diploid population \\
$V_{i,k}$ & gametic segregation variance associated with the production of a
  $k$-ploid gamete by individual $i$ \\
$\Vx_k$ & genetic variance associated with a haploid genome in the $k$-ploid
  reference population (i.e. a $k$-ploid non-inbred population at HWLE) \\
$\beta_{k}$ & scaling factor for allelic effects in $k$-ploids \\
$F_i$ & inbreeding coefficient in individual $i$ \\
$\Phi_{ij}$ & coancestry coefficient for individuals $i$ and $j$ \\
$\alpha_k$ & probability that the two genes at a locus in a diploid gamete
  formed by a $k$-ploid individual descend from the same parental gene copy\\
$\gamma$ & strength of directional selection in the new habitat\\
$\theta$ & trait value beyond which the growth rate becomes positive in the new
habitat \\ $w_{ij}$ & fitness of parental pair $(i,j)$ \\
$w_{ij}^{kl}$ & expected fitness of offspring from parental pair $(i,j)$ when
$i$ contributes a $k$-ploid gamete and $j$ contributes a $l$-ploid gamete \\
$\sigma_k$ & selfing rate in $k$-ploids \\
$\rho_k$ & probability of assortative mating in $k$-ploids \\
\cline{1-2}
\end{tabularx}%
\end{table}

@@

When two individuals mate, we assume they produce gametes according to their
ploidy level (\cref{eq:U}), which randomly combine to produce offspring (which
may be inviable if one of the contributing gametes is inviable).
Intrinsic fitness disadvantages associated with particular zygotic ploidy
levels or cross types (e.g. modeling phenomena such as `triploid block') can be
straighforwardly included at this level.
An analysis of a deterministic model (i.e. where $N \rightarrow \infty$) for
the cytotype dynamics and equilibrium cytotype composition under random mating
is included in \cref{s-sec:det} (see also \cite{felber1997,kauai2024}).
The stochastic version for finite and constant $N$ is analyzed briefly in
\cref{s-sec:mchain}.

@@

We model partial self-fertilization by assuming that a proportion
$\sigma_{c_i}$ of the ovules of individual $i$ with ploidy level $c_i$ are
fertilized by self-pollen, while the remaining proportion $1-\sigma_{c_i}$ are
fertilized by randomly sampled pollen (which may be self-pollen with
probability $1/N$). 
That is, the expected number of offspring from individual $i$ as mother
surviving after selection is
\begin{equation}
\Ex[w_i] = \sigma_{c_i} w_{ii} +
  (1-\sigma_{c_i})\left[\frac{1}{N}\sum_{j=1}^N w_{ij}\right]
\end{equation}
We hence assume no pollen limitation (all outcrossing ovules are fertilized),
and no pollen discounting (the probability of being a father is unaffected by
an individual's selfing rate).
When modeling self-incompatibility, we assume there is no intrinsic
disadvantage to self-incompatibility, except when there is only a single
individual in the population, i.e.
\begin{equation}
\Ex[w_i] = \begin{cases}
    \frac{1}{N-1}\sum_{j \ne i }w_{ij} & \text{if } N > 1\\ 
    0 & \text{if } N=1 \end{cases}
\end{equation}

We model assortative mating by ploidy level in a similar way, assuming that a
fraction $\rho_{c_i}$ of the ovules of individual $i$ are fertilized by pollen
sampled from the $c_i$-ploid portion of the population, while a fraction
$1-\rho_{c_i}$ is fertilized by pollen randomly sampled from the entire
population.
\begin{equation}
\Ex[w_i] = \rho_{c_i} \frac{1}{N_{c_{i}}} \sum_{j=1}^N \delta_{c_i,c_j}w_{ij}
 + (1-\rho_{c_i})
\left[\frac{1}{N}\sum_{j=1}^N w_{ij}\right]
\end{equation}

@@

Simulations for the mixed-ploidy model further confirm the correctness of our
infinitesimal approximation and highlight the importance of assumptions
regarding the scaling of allelic effects across ploidy levels
($\beta$ parameters, \cref{s-fig:vz}).
Although inbreeding is slower in autotetraploids than in diploids for the same
population size, the tetraploid fraction of a diploid-dominated mixed-ploidy 
population will have an equal or higher average inbreeding coefficient
(\cref{s-fig:f1}).
This is because in such a population, triploid and tetraploid individuals
mostly arise from gametes formed by diploid individuals or polyploid
individuals with very recent diploid ancestry (about $1+u+2v$ generations ago
for tetraploids, and $1+\frac{2}{3}(u+2v)$ generations ago for triploids, see
\cref{s-sec:ttdip}), so that the polyploid subpopulations will show an average
relatedness similar to that of the diploid population and not evolve as an
isolated higher-ploidy population would.
A nonzero probability of producing IBD diploid gametes ($\alpha_k > 0$) will
then further increase the inbreeding coefficient in the tetraploid and triploid
fraction of the population relative to their diploid progenitors
(\cref{s-fig:f1}).
Therefore, as long as diploids dominate, the effect of harboring some fraction
of the gene pool in polyploid individuals on the rate of inbreeding is
negligible, and we find that the evolution of the inbreeding coefficient over
time is well predicted by $1-e^{-t/2N_e}$, where the inbreeding-effective
population size is, to first order in $u$, given by $(1-2u)N$
(\cref{s-sec:effsize}).

@@

\begin{figure}[t]
\centering
\includegraphics[width=\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/est1b.pdf}
\caption{
(A) Probability of establishment from a single diploid or tetraploid individual
with trait value $z=0$ for increasing selection intensity $\gamma$.
We assume $m=0$ and $u=0$, i.e. there is no migration, and no unreduced
gametes are produced. The trait is scaled in tetraploids so as to yield the
same genetic variance at HWLE ($\beta_4 = \sqrt{1/2}$)
(B) Probability of a tetraploid individual with trait value $z = 0$
successfully founding a population ($P_4$), relative to the probability for a
diploid individual with the same trait value ($P_2$).
The vertical dashed line marks $\beta_4=\sqrt{1/2}$, for which the variance at
HWLE is identical between diploids and autotetraploids.
(C) Probability of tetraploid establishment relative to the probability of
diploid establishment (on a $\log_{10}$ scale) across a range of values for
$\gamma$ and $\beta_4$. The vertical line again marks $\beta_4=\sqrt{1/2}$.
All results are estimated from 500.000 replicate simulations and assume
$\theta=2$.
\label{fig:est1}}
\end{figure}

@@

We first consider the establishment of a population from a single migrant
individual with trait value $z_0 = 0$.
We assume $u=0$ and compare the probability of establishment when the migrant
is diploid vs. tetraploid (\cref{fig:est1}).
As noted by \cite{barton2018}, the establishment probability (for a given
cytotype) depends essentially on two dimensionless parameters,
$\gamma\sqrt{2V}$ and $\theta/\sqrt{2V}$, corresponding to the intensity of
selection and the degree of maladaptation, respectively.
We shall scale our results accordingly, assuming $2V=1$ throughout, and shall
assume $\theta=2$, i.e. adaptation has to increase the trait value by two
standard deviations to achieve a positive growth rate.

We find that reduced inbreeding in tetraploids substantially increases the
establishment probability.
Indeed, in the case where allelic effects are scaled across ploidy levels
to yield the same equilibrium genetic variance ($\beta = \sqrt{1/2}$, so that
polyploid migrants are not more likely to have high fitness than diploid
migrants), the establishment probability for tetraploids can be almost five
times as high as in diploids depending on the selection gradient
(\cref{fig:est1}A).
As the segregation variance and initial trait value are the same across these
simulations, this is a consequence of the reduced rate of inbreeding, which
slows down the exhaustion of the genetic variance carried by the initial
migrant individual.

While the probability of establishment becomes lower as the strength of
selection becomes large, the establishment probability does not decrease
monotonically with $\gamma$, i.e. as selection becomes very weak, the
establishment probability also decreases.
This happens because, although the probability of surviving the first couple of
generations becomes higher when selection is weaker, adaptation (i.e. increase
in the trait mean) will be slower,
and the risk that genetic variation is exhausted due to inbreeding before the
population is able to reach a consistently positive growth rate is increased as
a result.

Evidently, the scaling of the genetic variance across ploidy levels has a
profound effect on the establishment probability, but only when $\beta$ is
close to 0.5 (i.e. individual alleles have almost half the effect size in
tetraploids compared to diploids) is the benefit of the slower rate of
inbreeding in tetraploids canceled (\cref{fig:est1}B,C).

@@

\begin{figure}[t]
\centering
\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration1b.pdf}
\caption{
Establishment with recurrent migration.
(A) Expected time until a population is established in the novel habitat for
increasing rates of migration and different degrees of maladaptation
($\bar{z}_s$, the mean trait value in the source population). 
(B) Proportion of simulation replicates in which tetraploids established. 
The dashed horizontal line marks the proportion of tetraploid migrants (i.e.
the proportion of tetraploids at equilibrium in the source population).
The solid horizontal lines mark $\frac{p_4\pi_4}{(p_4\pi_4 + p_2\pi_2)}$, where $p_2$ and
$p_4$ are the establishment probabilities for a single diploid and a single
tetraploid individual in the absence of migration (as in \cref{fig:est1}, but
with unreduced gamete formation included).
All results are based on 10.000 replicate simulations.
We assume $\gamma=0.25, \theta=1.5$ and $u=v=0.05$.
\label{fig:estwmig}}
\end{figure}

We next consider establishment in the new habitat when there is a continuous
influx of migrants (at rate $m$) coming from a large, noninbred and
predominantly diploid source population at cytotype equilibrium.
In this setting, establishment is certain to happen eventually, and we are
interested in the probability that a tetraploid population establishes before a
diploid one does, as well as the effects of maladaptive gene flow and MCE
(which both increase with increasing migration rate) on the time until
establishment.

We hypothesized that two counteracting processes affect the probability of
autotetraploid establishment in this scenario.
On the one hand, increased migration will increase the probability that an
otherwise likely successful tetraploid migrant suffers from MCE in the early
generations while the population size is low, because migrants are likely to be
diploid.
On the other hand, tetraploids are strongly reproductively isolated from
a typical migrant, so that they are less prone to maladaptive swamping.
Hence, conditional on evading MCE, they should be able to adapt to the new
habitat at a rate which is not strongly affected by the migration rate.
This contrasts with diploids, which interbreed freely with maladapted
migrants, resulting in a pulling back of the trait mean towards that of the
source population.

As expected, we find that the time to establishment (of a population of either
ploidy level) first decreases with increasing migration as a result of a larger
influx of potentially succesful migrants, but later increases with increasing
migration as a result of swamping by gene flow (\cref{fig:estwmig}A).
The probability that a tetraploid population establishes decreases
monotonically with increasing migration and is considerably larger than the
probability that a migrant individual is tetraploid over a large part of the
parameter range (\cref{fig:estwmig}B).
Together, this suggests that the negative effects of MCE generally outweigh the
benefits that come from reproductive isolation as migration increases.

\subsection*{Loss of self-incompatibility and assortative mating}

When polyploidization disrupts an existing SI system, we expect that
tetraploids suffer less from MCE as some portion of their ovules are now
assured to be fertilized by diploid gametes.
At the same time, we expect that accelerated inbreeding in selfing tetraploids
diminishes the adaptive advantage of tetraploids that we found above.
As shown in \cref{fig:selfing}, we find that when polyploidization is
associated with the loss of a SI system, tetraploids have a strongly increases
establishment probability.
This is the case even when the selfing rate $\sigma$ in tetraploids is zero (in
which case, under our modeling assumptions, there is only random selfing, i.e.
the \textit{realized} selfing rate in tetraploids is $1/N$).
Besides reducing MCE, selfing is also expected to simultaneously reduce the
effects of maladaptive gene flow, and we indeed find that when the selfing rate
is sufficiently high ($\ge 0.4$ in \cref{fig:selfing}), the relative
establishment probability of tetraploids increases with increasing migration
rate.
In this regime, the increased strength of MCE due to increased migration is
compensated by the stronger maladaptive gene flow experienced by diploids.

\begin{figure}
\centering
\includegraphics[width=0.9\linewidth]{/home/arthur_z/dev/InfGenetics/doc/img/estwmigration-si2.pdf}
\caption{
Establishment with recurrent migration and loss of self-incompatibility in
polyploids (diploids are assumed to be self-incompatible).
(A) Expected time until a population is established in the novel habitat for
increasing rates of migration and different self-fertilization rates in
polyploids. Triploids and tetraploids have the same selfing rate. 
Diploids are assumed to be self-incompatible.  Note that $\sigma=0.0$ refers to
random self-fertilization (i.e.  self-fertilization with probability $1/N$). 
(B) Proportion of simulation replicates in which tetraploids established. 
All results are based on 25.000 replicate simulations.
We assume $\gamma=0.25, \theta=1.5$ and $u=v=0.05$.
\label{fig:selfing}}
\end{figure}
